---
title: "BC Analysis"
author: "Maggie Li (ml4424)"
date: "4/8/2021"
output: html_document
---

```{r load libraries}
library(tidyverse)
library(janitor)
library(stats)
library(lme4)
library(raster)
library(sf)
library(usmap)
library(tictoc)
require(lmerTest)
library(ncdf4)
```


# Aim 1: Compare county-level PM2.5 component concentrations between AI and non-AI-populated counties

```{r read in file with county type defined for each 3109 study counties}
ai_county_fips = read_csv("data/ai_counties.csv") 

nai_county_fips = read_csv("data/nai_counties.csv") 
```

## Extract county-level BC concentrations for 2000-2017

Data Source: Randall Martin's PM2.5 Model https://sites.wustl.edu/acag/datasets/surface-pm2-5/ 
Note: The raw data show the proportion of each component, so have to multiply it by the total PM concentration that year

For Vivian: The next 2 code chunks will take a while to run, especially looping through all years. If it's too much, you can run just one year or a few; the outputted datasets are already in the folders indicated by write_csv.

```{r read and extract county level BC for 2000 2001 2017 only}
# read in all counties shapefile
counties_shp <- "data/cb_2018_us_county_500k/cb_2018_us_county_500k.shp"
all_counties <- st_read(counties_shp, stringsAsFactors = FALSE)

# group all_counties by state fips, exclude territories and hawaii and alaska
all_counties <- all_counties %>% arrange(STATEFP) %>%
  filter(STATEFP != "02",
         STATEFP != "15",
         STATEFP <= "56")
unique(all_counties$STATEFP)
all_counties

# each state's fips code in a vector
state_fips <- fips(state.name, county = c())

# remove hawaii and alaska (FIPS are 15 and 02)
state_fips <- state_fips[!(state_fips %in% c("02","15"))]

# save each state's counties as separate sf's, in a list. the function below will iterate through all of these states.
state_list <- list(length(state_fips)) # create list of length 48
for (i in 1:length(state_fips)){
  state_list[[i]] <- all_counties %>%
    filter(STATEFP==state_fips[i])
}
# state_list # each list value = one state; includes polygon counties in that state; 48 states total

# read in ncdf as raster stack for BC in 2000, 2017
bc_2000 = raster("data/raw_components/bc/bc_2000.nc")
bc_2017 = raster("data/raw_components/bc/bc_2017.nc")
bc_2000 = stack(bc_2000)
bc_2017 = stack(bc_2017)

# read in ncdf as raster stack for total PM in 2000, 2017
pm_2000 = raster("data/raw_components/total_pm/pm_2000.nc")
pm_2017 = raster("data/raw_components/total_pm/pm_2017.nc")
pm_2000 = stack(pm_2000)
pm_2017 = stack(pm_2017)

# get the actual raster matrix for BC through multiplying percentage with total PM
bc_actual_2000 = (bc_2000*pm_2000)/100
mean(getValues(bc_actual_2000), na.rm = T)

bc_actual_2017 = (bc_2017*pm_2017)/100
mean(getValues(bc_actual_2017), na.rm = T)

# create empty lists to save dfs into
bc_2000_list = list()
bc_2017_list = list()

# view values for 2017
raster::extract(bc_2017, state_list[[1]], fun=mean, na.rm=TRUE, df=TRUE)
mean(getValues(bc_2017), na.rm = T)

# write loop to extract county BC levels for 2000: this will take a while to run

tic() # tictoc helps record the time it takes to run things in R
for (i in 1:48){ # 48 states total FIPS = 01 (AL), 04 (AZ)
  bc_2000_list[[i]] <- raster::extract(bc_actual_2000,
                                       state_list[[i]],
                                       fun=mean, na.rm=TRUE, df=TRUE) #specify function = mean to extract mean concentrations
  bc_2000_list[[i]]$County <- state_list[[i]]$COUNTYFP
  bc_2000_list[[i]]$State <- state_fips[i]
  bc_2000_list[[i]]$FIPS <- paste(bc_2000_list[[i]]$State,
                                  bc_2000_list[[i]]$County)
  bc_2000_list[[i]]$FIPS <- str_replace_all(bc_2000_list[[i]]$FIPS, " ", "")}

toc()

# PM25 for each county should be saved now in the previously empty list
# (each list item = one state, each row = county within that state)
bc_2000_list[[1]]
bc_2000_df = do.call(rbind, bc_2000_list)
# write out dataset to intermediate comp_concentrations folder
write_csv(bc_2000_df,
          "data/comp_concentrations/bc/bc_2000_df.csv")

# read in ncdf as raster stack for BC in 2001
bc_2001 = raster("data/raw_components/bc/bc_2001.nc")
bc_2001 = stack(bc_2001)


# read in ncdf as raster stack for total PM in 2001
pm_2001 = raster("data/raw_components/total_pm/pm_2001.nc")
pm_2001 = stack(pm_2001)

# get the actual raster matrix for BC through multiplying percentage with total PM
bc_actual_2001 = (bc_2001*pm_2001)/100
mean(getValues(bc_actual_2001), na.rm = T) # average county mean

# create empty lists to save dfs into
bc_2001_list = list()

# write loop to extract county BC levels for 2001: this will take a while to run

tic() # tictoc helps record the time it takes to run things in R
for (i in 1:48){ # 48 states total FIPS = 01 (AL), 04 (AZ)
  bc_2001_list[[i]] <- raster::extract(bc_actual_2001,
                                       state_list[[i]],
                                       fun=mean, na.rm=TRUE, df=TRUE) #specify function = mean to extract mean concentrations
  bc_2001_list[[i]]$County <- state_list[[i]]$COUNTYFP
  bc_2001_list[[i]]$State <- state_fips[i]
  bc_2001_list[[i]]$FIPS <- paste(bc_2001_list[[i]]$State,
                                  bc_2001_list[[i]]$County)
  bc_2001_list[[i]]$FIPS <- str_replace_all(bc_2001_list[[i]]$FIPS, " ", "")}

toc()

# PM25 for each county should be saved now in the previously empty list
# (each list item = one state, each row = county within that state)
bc_2001_list[[1]]
bc_2001_df = do.call(rbind, bc_2001_list)
write_csv(bc_2001_df,
          "data/comp_concentrations/bc/bc_2001_df.csv")

# write loop to extract county BC levels for 2017: this will take a while to run

tic() # tictoc helps record the time it takes to run things in R
for (i in 1:48){ # 48 states total FIPS = 01 (AL), 04 (AZ)
  bc_2017_list[[i]] <- raster::extract(bc_actual_2017,
                                       state_list[[i]],
                                       fun=mean, na.rm=TRUE, df=TRUE) #specify function = mean to extract mean concentrations
  bc_2017_list[[i]]$County <- state_list[[i]]$COUNTYFP
  bc_2017_list[[i]]$State <- state_fips[i]
  bc_2017_list[[i]]$FIPS <- paste(bc_2017_list[[i]]$State,
                                  bc_2017_list[[i]]$County)
  bc_2017_list[[i]]$FIPS <- str_replace_all(bc_2017_list[[i]]$FIPS, " ", "")}

toc()

# PM25 for each county should be saved now in the previously empty list
# (each list item = one state, each row = county within that state)
bc_2017_list[[1]]
bc_2017_df = do.call(rbind, bc_2017_list)
write_csv(bc_2017_df,
          "data/bc_2017_df.csv")
```

```{r run loop to read in and get BC data all years}
# Loop from 2001-2017
bc_years = c("bc_2000", "bc_2001","bc_2002", "bc_2003", "bc_2004", "bc_2005",
             "bc_2006", "bc_2007", "bc_2008", "bc_2009",
             "bc_2010", "bc_2011", "bc_2012", "bc_2013",
             "bc_2014", "bc_2015", "bc_2016", "bc_2017")

pm_years = c("pm_2000", "pm_2001", "pm_2002", "pm_2003", "pm_2004", "pm_2005",
             "pm_2006", "pm_2007", "pm_2008", "pm_2009",
             "pm_2010", "pm_2011", "pm_2012", "pm_2013",
             "pm_2014", "pm_2015", "pm_2016", "pm_2017")

# Extract Data for DC
## Note: we have to extract concentrations in DC separately,  because DC was not a state in the state_list sf file.
## create a df just DC all years, join with the existing components_year_df files in data/county_concentrations
DC_state_list <- all_counties %>%
    filter(STATEFP== "11")

# create empty list to save DC PM data into; (each list element is 1 df with one row, for the DC estimate that year.)
bc_dc_list = list()

# count variable for year to name saved out df
year = 2000
## Run loop for 2000 through 2017
for (i in 1:18){
  # read in ncdf as raster stack for bcs
  bc_year = raster(paste("data/raw_components/bc/", bc_years[i], ".nc", sep = ""))
  bc_year = stack(bc_year)


  # read in ncdf as raster stack for total PM in 2001
  pm_year = raster(paste("data/raw_components/total_pm/", pm_years[i], ".nc", sep = ""))
  pm_year = stack(pm_year)

  # get the actual raster matrix for BC through multiplying percentage with total PM
  bc_actual_year = (bc_year*pm_year)/100
  mean(getValues(bc_actual_year), na.rm = T) # average county mean

# extract data for just DC (fips = 11)
  bc_dc_list[[i]] <- raster::extract(bc_actual_year,
                            DC_state_list,
                            fun=mean, na.rm=TRUE, df=TRUE) %>% # specify function = mean to extract mean concentrations
    mutate(year = year) %>%
    dplyr::rename(FIPS = ID) %>%
    dplyr::rename(bc = layer) %>%
    mutate(State = 11) %>%
    mutate(FIPS = 11001)
  bc_dc_list[[i]]$County <- "001"
  year = year + 1


}
toc()
bc_dc_list
bc_dc = do.call(rbind, bc_dc_list)
bc_dc


tic()

## count variable for year to name saved out df
year = 2000

## Run loop for 2000 through 2017
for (i in 1:18){
  # read in ncdf as raster stack for bcs
  bc_year = raster(paste("data/raw_components/bc/", bc_years[i], ".nc", sep = ""))
  bc_year = stack(bc_year)


  # read in ncdf as raster stack for total PM in 2001
  pm_year = raster(paste("data/raw_components/total_pm/", pm_years[i], ".nc", sep = ""))
  pm_year = stack(pm_year)

  # get the actual raster matrix for BC through multiplying percentage with total PM
  bc_actual_year = (bc_year*pm_year)/100
  mean(getValues(bc_actual_year), na.rm = T) # average county mean

  # create empty lists to save dfs into
  bc_year_list = list()

  # write loop to extract county BC levels for all states in given year: this will take a while to run

  tic() # tictoc helps record the time it takes to run things in R
  for (i in 1:48){ # 48 contiguous states total FIPS = 01 (AL), 04 (AZ)
    bc_year_list[[i]] <- raster::extract(bc_actual_year,
                                         state_list[[i]],
                                         fun=mean, na.rm=TRUE, df=TRUE) #specify function = mean to extract mean concentrations
    bc_year_list[[i]]$County <- state_list[[i]]$COUNTYFP
    bc_year_list[[i]]$State <- state_fips[i]
    bc_year_list[[i]]$FIPS <- paste(bc_year_list[[i]]$State,
                                    bc_year_list[[i]]$County)
    bc_year_list[[i]]$FIPS <- str_replace_all(bc_year_list[[i]]$FIPS, " ", "")}

  toc()

  # PM25 for each county should be saved now in the previously empty list
  # (each list item = one state, each row = county within that state)
  bc_year_df = do.call(rbind, bc_year_list)
  write_csv(bc_year_df,
            paste("data/county_concentrations/bc/bc_", year, "_df.csv", sep = ""))
  year = year + 1
}
toc()
```


```{r combine all years into one df}
# Read in model data into list of length 19 (# study years)
bc_allyrs <- list()
yr = 2000
for (i in 1:18){
  bc_allyrs[[i]] <- read_csv(file = paste('data/county_concentrations/bc/bc_',
                                           yr, '_df.csv', sep = '')) %>%
    dplyr::rename(bc = layer) %>% # rename "layer" column to specify component (BC)
    mutate(year = yr) # add column for year
  yr <- yr + 1
}

bc_allyrs = do.call(rbind, bc_allyrs) # join list into one df

# UPDATE: join with DC data
bc_allyrs = bc_allyrs %>% dplyr::select(-ID)
bc_allyrs = rbind(bc_allyrs, bc_dc)

summary(bc_allyrs)
length(unique(bc_allyrs$FIPS))

# write out Final joined dataset w/ 48 states + DC
write_csv(bc_allyrs, "data/county_concentrations/allyrs/bc_allyrs.csv")

```

## Assign BC estimates to AI vs non-AI counties-- 

Note to Vivian: You can use the pre-generated data (since the 3 code chunks above extract the county-level estimates)

```{r join layer to AI county}
bc_allyrs = read_csv("data/county_concentrations/allyrs/bc_allyrs.csv")

all_county_types = rbind(ai_county_fips, nai_county_fips) %>% 
  rename(FIPS = County)

bc_ai_join = left_join(bc_allyrs, all_county_types) %>% 
  dplyr::select(FIPS, bc, year, county_type) %>% 
  rename(County = FIPS) %>% 
  mutate(county_type = replace_na(county_type, 1)) ## county FIPS 46102 is showing up as an NA
  
bc_ai_join

```

```{r join with covariates}
covariates = read_csv("data/covariates.csv")
summary(covariates)
bc_ai_join = bc_ai_join %>% 
  left_join(covariates)
bc_ai_join

# Compare summaries in all counties, AI, non-AI
summary(bc_ai_join)
summary(bc_ai_join %>% filter(county_type == 1))
summary(bc_ai_join %>% filter(county_type == 0))

# note higher median/mean levels in nAI counties (unadjusted)
```

```{r add columns for deciles}
# Split population density and hhincome into deciles for model
bc_ai_join$popd_q <- cut(bc_ai_join$pop_density, 
                         quantile(bc_ai_join$pop_density, seq(0,1,0.1)), 
                         include.lowest = TRUE)
bc_ai_join$hhinc_q <- cut(bc_ai_join$hh_income, 
                          quantile(bc_ai_join$hh_income, seq(0,1,0.1)), 
                          include.lowest = TRUE)
sum(table(bc_ai_join$popd_q, exclude = NULL)) == dim(bc_ai_join)[1]
bc_ai_join
```
```{r boxplot of BC distribution by AI county type}
## distribution of BC by county type
boxplot(bc~county_type,data=bc_ai_join, main="BC by AI-pop county type",
   xlab="County Type", ylab="BC (μg/m3)")

  # set referent for df
bc_ai_join = bc_ai_join %>% 
  mutate(county_type = factor(county_type)) %>% 
  mutate(county_type = relevel(county_type, ref="0"))

## Save out dataframe 
write_csv(bc_ai_join,
          "data/county_concentrations/joined_dta/bc_ai_join.csv")
```

## Run statistical models;
```{r Run lmer models}
# unadjusted
bc_unadj = lmer(bc ~ county_type + as.factor(year) +
                    (1|State/County),
                  data = bc_ai_join)

## 95% CI: point estimate, LL, UL
paste(summary(bc_unadj)$coefficients[2,1] %>% round(digits = 3),
      " (",
      round(summary(bc_unadj)$coefficients[2,1] - 1.96*summary(bc_unadj)$coefficients[2,2], digits = 3),
      ", ",
      round(summary(bc_unadj)$coefficients[2,1] + 1.96*summary(bc_unadj)$coefficients[2,2], digits = 3),
      ")", sep = "")
paste(summary(bc_unadj)$coefficients[2,1] %>% round(digits = 2),
      " (",
      round(summary(bc_unadj)$coefficients[2,1] - 1.96*summary(bc_unadj)$coefficients[2,2], digits = 2),
      ", ",
      round(summary(bc_unadj)$coefficients[2,1] + 1.96*summary(bc_unadj)$coefficients[2,2], digits = 2),
      ")", sep = "")

# additionally adjusted for popd
bc_partial = lmer(bc ~ county_type + as.factor(year) +
                  popd_q +
                  (1|State/County),
                data = bc_ai_join)

## 95% CI: point estimate, LL, UL
paste(summary(bc_partial)$coefficients[2,1] %>% round(digits = 3),
      " (",
      round(summary(bc_partial)$coefficients[2,1] - 1.96*summary(bc_partial)$coefficients[2,2], digits = 3),
      ", ",
      round(summary(bc_partial)$coefficients[2,1] + 1.96*summary(bc_partial)$coefficients[2,2], digits = 3),
      ")", sep = "")
paste(summary(bc_partial)$coefficients[2,1] %>% round(digits = 2),
      " (",
      round(summary(bc_partial)$coefficients[2,1] - 1.96*summary(bc_partial)$coefficients[2,2], digits = 2),
      ", ",
      round(summary(bc_partial)$coefficients[2,1] + 1.96*summary(bc_partial)$coefficients[2,2], digits = 2),
      ")", sep = "")

# full adj model
bc_full = lmer(bc ~ county_type + as.factor(year) +
                    popd_q + 
                    hhinc_q +
                    (1|State/County),
                  data = bc_ai_join, REML=FALSE)

## 95% CI: point estimate, LL, UL
paste(summary(bc_full)$coefficients[2,1] %>% round(digits = 3),
      " (",
      round(summary(bc_full)$coefficients[2,1] - 1.96*summary(bc_full)$coefficients[2,2], digits = 3),
      ", ",
      round(summary(bc_full)$coefficients[2,1] + 1.96*summary(bc_full)$coefficients[2,2], digits = 3),
      ")", sep = "")
paste(summary(bc_full)$coefficients[2,1] %>% round(digits = 2),
      " (",
      round(summary(bc_full)$coefficients[2,1] - 1.96*summary(bc_full)$coefficients[2,2], digits = 2),
      ", ",
      round(summary(bc_full)$coefficients[2,1] + 1.96*summary(bc_full)$coefficients[2,2], digits = 2),
      ")", sep = "")

# fully adjusted with interaction term for year
bc_full_interx = lmer(bc ~ county_type + as.factor(year) +
                    popd_q + 
                    hhinc_q +
                    county_type*as.factor(year) +
                    (1|State/County),
                  data = bc_ai_join, REML=FALSE)
summary(bc_full_interx)
```

```{r Save out results from main effects only model and interx effects model}

## save model output for main effects only model
saveRDS(bc_full,
        file = "intermediate/model_outputs/main_results/bc_main_effects.rds")
## save model output for main effects only model
saveRDS(bc_full_interx,
        file = "intermediate/model_outputs/main_results/bc_interx.rds")

```


```{r InterX plot from lmer}
# create vcov matrix of main effects and interX

# extract vcov matrix of the county type main effect and interaction effect for each categorical year
native_yr_vcov <- vcov(bc_full_interx)[c(2,seq(38,54)), c(2,seq(38,54))]

# calculate all the variances for all the years i.e. var(x+y); should be 19 total entries --> var_vector
# var_vector should be the same repeating value (from the diagonal of the vcov matrix above)
var_vector = c()
for (i in 2:18){
  var_vector[1] <- native_yr_vcov[1,1]
  var_vector[i] <- native_yr_vcov[1,1] + native_yr_vcov[i,i] + 2 * native_yr_vcov[i,1]
}
var_vector
sd_vector <- sqrt(var_vector)
length(sd_vector)
#matrix with 19 cols for 19 years, three rows: one for effect estimate of total effect per year (total effect = main effect + interx effect), one for CI lower, one for CI upper
bc_decline <- data.frame()
bc_decline[1,1] <- summary(bc_full_interx)$coefficients[2,1]
bc_decline[1,2] <- summary(bc_full_interx)$coefficients[2,1] - 1.96*sd_vector[1]
bc_decline[1,3] <- summary(bc_full_interx)$coefficients[2,1] + 1.96*sd_vector[1]

# fill in matrix thru loop for every following year
yr_ct <- 38
for (i in 2:18){
  bc_decline[i,1] <- summary(bc_full_interx)$coefficients[2,1]+
    summary(bc_full_interx)$coefficients[yr_ct,1]
  
  bc_decline[i,2] <- summary(bc_full_interx)$coefficients[2,1]+
    summary(bc_full_interx)$coefficients[yr_ct,1] - 1.96*sd_vector[i]
  
  bc_decline[i,3] <- summary(bc_full_interx)$coefficients[2,1]+
    summary(bc_full_interx)$coefficients[yr_ct,1] + 1.96*sd_vector[i]
  yr_ct <- yr_ct + 1
}

bc_decline

colnames(bc_decline) <- c('estimate', 'cl_lower', 'cl_upper') # set col names

bc_decline$year <- seq(2000, 2017) # column for year


#PLOT OF TOTAL EFFECT OF NATIVE OVER TIME; updated 12/14 to be consistent with Figure 3 layout (slanted x-axis label)
bc_interx_plot <- ggplot() + 
  theme_linedraw() + 
  geom_line(data = bc_decline,
            aes(x=year, y = estimate)) +
  geom_line(data=bc_decline,
            aes(x=year, y=cl_lower), linetype = "dashed") +
  geom_line(data=bc_decline,
            aes(x=year, y=cl_upper), linetype = "dashed") +
  ylim(-0.6, 0.6) +
  labs(x = "Year",
       y = expression(paste("Mean Difference in BC (", mu, "g/", m^3, ")")),
       fill = "County Type") +
  theme(plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 12),
        axis.title.x=element_blank(),
        panel.background = element_rect(fill='transparent'), 
        plot.background = element_rect(fill='transparent', color=NA)) +
  scale_x_continuous(breaks = seq(2000,2017,1), expand = c(0, 0)) +
  guides(x =  guide_axis(angle = 45)) + 
  geom_hline(yintercept=0, linetype="solid", color = "red")
bc_interx_plot
ggsave("figures/effectmod_plots/main/bc.png", width = 6, height = 4, dpi = 300,
       bg='transparent')

```

To Vivian: We also tried running smoothed BC trends over time by county type (https://fromthebottomoftheheap.net/2017/10/10/difference-splines-i/)

Given similarities with the linear models, we did not choose to include them in our paper, opting instead for the GLMM results above. You can run the below to check it out if you wish to.

```{r Smooth interX model with time 10 knots}
# using splines
library(splines)
library(mgcv)
library(gamm4)
library(reshape2)
bc_full_interx_smooth = gamm4(bc ~ s(year, by = county_type, k = 10) + # specify k knots
                    county_type + 
                      popd_q + 
                    hhinc_q,
                    random = ~(1|State/County),
                  data =  bc_ai_join)
summary(bc_full_interx_smooth)
summary(bc_full_interx_smooth$gam)

# view smooth plots for county type = 0 and 1
plot(bc_full_interx_smooth$gam, pages = 1)

# empty df for predicted values from gam; decile of popd and hhinc correspond to most common quantile for non-AI pop counties 
pdat = expand.grid(year = seq(2000, 2017, length = 170), # length = 170 predicts 10 values per year (200-2016)
                    county_type = c(0, 1),
                   popd_q = "(160,382]",
                   hhinc_q = "(5.75e+04,1.16e+05]")

# add in predicted values, get difference for each year and plot; is ggplot smooth adequate?
pdat %>%
  modelr::add_predictions(bc_full_interx_smooth$gam) %>%
  dcast(year ~ county_type, value.var = "pred", fill = 0) %>%
  mutate(diff = `1` - `0`) %>%
  ggplot(aes(x = year, y = diff)) +
           geom_line()

# add in predicted values from gam
xp <- predict(bc_full_interx_smooth$gam, newdata = pdat, type = 'lpmatrix')
xp

## which cols of xp relate to splines of interest?
c1 <- grepl('1', colnames(xp))
c2 <- grepl('0', colnames(xp))
## which rows of xp relate to sites of interest?
r1 <- with(pdat, county_type == '1')
r2 <- with(pdat, county_type == '0')

## difference rows of xp for data from comparison
X <- xp[r1, ] - xp[r2, ]
## zero out cols of X related to splines for other lochs
X[, ! (c1 | c2)] <- 0
## zero out the parametric cols
X[, !grepl('^s\\(', colnames(xp))] <- 0
#extract betas
dif <- X %*% coef(bc_full_interx_smooth$gam)
#extract se
se <- sqrt(rowSums((X %*% vcov(bc_full_interx_smooth$gam)) * X))
#calculate CI
crit <- qt(.975, df.residual(bc_full_interx_smooth$gam))
upr <- dif + (crit * se)
lwr <- dif - (crit * se)

# function to do it all in 1
smooth_diff <- function(model, newdata, f1, f2, var, alpha = 0.05,
                        unconditional = FALSE) {
    xp <- predict(model, newdata = newdata, type = 'lpmatrix')
    c1 <- grepl(f1, colnames(xp))
    c2 <- grepl(f2, colnames(xp))
    r1 <- newdata[[var]] == f1
    r2 <- newdata[[var]] == f2
    ## difference rows of xp for data from comparison
    X <- xp[r1, ] - xp[r2, ]
    ## zero out cols of X related to splines for other lochs
    X[, ! (c1 | c2)] <- 0
    ## zero out the parametric cols
    X[, !grepl('^s\\(', colnames(xp))] <- 0
    dif <- X %*% coef(model)
    se <- sqrt(rowSums((X %*% vcov(model, unconditional = unconditional)) * X))
    crit <- qt(alpha/2, df.residual(model), lower.tail = FALSE)
    upr <- dif + (crit * se)
    lwr <- dif - (crit * se)
    data.frame(pair = paste(f1, f2, sep = '-'),
               diff = dif,
               se = se,
               upper = upr,
               lower = lwr)
}

comp1 <- smooth_diff(bc_full_interx_smooth$gam, pdat, '1', '0', 'county_type')
comp <- cbind(year = seq(2000, 2017, length = 170),
              rbind(comp1))

comp

# plot smoothed difference
ggplot(comp, aes(x = year, y = diff, group = pair)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
    geom_line() +
  labs(x = "Year",
       y = expression(paste("Mean Difference in BC (", mu, "g/", m^3, ")")),
       fill = "County Type") +
  theme(plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 12),
        axis.title.x=element_blank()) +
  scale_x_continuous(breaks = seq(2000,2017,1), expand = c(0, 0)) +
  guides(x =  guide_axis(angle = 45)) +
  ylim(-0.11, 0.1) +
  geom_hline(yintercept=0, linetype="solid", color = "red") +
  theme_linedraw()

# Plot difference in splines by county type using add_predictions only
pdat %>%
  modelr::add_predictions(bc_full_interx_smooth$gam) %>%
  dcast(year ~ county_type, value.var = "pred", fill = 0) %>%
  mutate(diff = `1` - `0`) %>%
  ggplot(aes(x = year, y = diff)) +
           geom_line()
ggsave("figures/effectmod_plots/old/bc/bc_smooth.png", width = 6, height = 4, dpi = 300)

# compare with original glmm
bc_interx_plot

## Compare fitted values between GLMM and GAMM

tibble(
  fit_glmm = fitted(bc_full_interx),
  fit_gamm = fitted(bc_full_interx_smooth$gam)) %>% 
  ggplot(aes(x = fit_glmm, y = fit_gamm)) + 
  geom_point() 

tibble(
  fit_glmm = fitted(bc_full_interx),
  fit_gamm = fitted(bc_full_interx_smooth$gam)) %>% 
  mutate(diff = fit_glmm - fit_gamm) %>% 
  ggplot(aes(x = diff)) + geom_histogram() # 


length(fitted(bc_full_interx))
length(fitted(bc_full_interx_smooth$gam))
complete.cases(bc_ai_join)

```

Next steps: We wanted to check if glmm and gamm predict the same way (include random effects or just fixed effects). We check glm vs. gam fitted values, no random effects (just year), then just year and county as a fixed effect (not random effect), then year and county as a random effect. We also compare models with fixed vs random county effects. We didn't find substantial differences depending on the diff packages used.

```{r Compare GLM and GAM}
# # 1. run GLM and GAM with only year as fixed effect
# bc_glm = glm(bc ~ as.factor(year),
#     data = bc_ai_join)
# summary(bc_glm)
# bc_gam = gamm4(bc ~ s(year),
#     data = bc_ai_join)
# summary(bc_gam)
# 
# ## compare GLM and GAM fitted values
# glm_gam_pt = tibble(
#   fit_glm = fitted(bc_glm),
#   fit_gam = fitted(bc_gam$gam)) %>% 
#   ggplot(aes(x = fit_glm, y = fit_gam)) + 
#   geom_point() 
# 
# glm_gam_hist = tibble(
#   fit_glm = fitted(bc_glm),
#   fit_gam = fitted(bc_gam$gam)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# require(cowplot)
# plot_grid(glm_gam_pt, glm_gam_hist, ncol=2)
# ### notes: seem to fit along the identity line (at least, not the horizontal striations we saw above). small range of differences in values based on histogram.
# 
# # 2. run GLM and GAM with year AND county as fixed effects
# ## sample 10,000 rows because otherwise it will take forever to run
# 
# bc_ai_sample = sample_n(bc_ai_join, 10000)
# bc_glm_countyfixed = glm(bc ~ as.factor(year) + County,
#     data = bc_ai_sample)
# summary(bc_glm_countyfixed)
# bc_gam_countyfixed = gamm4(bc ~ s(year) + County,
#     data = bc_ai_sample)
# summary(bc_gam_countyfixed)
# 
# ## compare GLM and GAM fitted values
# glm_gam_county_pt = tibble(
#   fit_glm = predict(bc_glm_countyfixed),
#   fit_gam = predict(bc_gam_countyfixed$gam)) %>% 
#   ggplot(aes(x = fit_glm, y = fit_gam)) + 
#   geom_point() 
# 
# glm_gam_county_hist = tibble(
#   fit_glm = predict(bc_glm_countyfixed),
#   fit_gam = predict(bc_gam_countyfixed$gam)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# 
# 
# plot_grid(glm_gam_county_pt, glm_gam_county_hist)
# fixef(bc_gam_countyfixed$mer)
# ### notes: also seem to fit along the identity line (at least, not the horizontal striations we saw above). small range of differences in values based on histogram. a good sign!
# 
# 
# # 3. Run GLMM and GAMM with county as a nested random effect within states
# ## use sampled 10,000 rows because otherwise it will take forever to run
# bc_glmm_county = lmer(bc ~ as.factor(year) + (1|State/County),
#     data = bc_ai_join)
# summary(bc_glmm_county)
# bc_gamm_county = gamm4(bc ~ s(year, k = 10),
#                            random = ~(1|State/County),
#                        data = bc_ai_join)
# summary(bc_gamm_county$gam)
# 
# ## compare GLMM and GAMM fitted values
# glmm_gamm_pt = tibble(
#   fit_glmm = fitted(bc_glmm_county),
#   fit_gamm = fitted(bc_gamm_county$mer)) %>% 
#   ggplot(aes(x = fit_glmm, y = fit_gamm)) + 
#   geom_point() 
# 
# glmm_gamm_hist = tibble(
#   fit_glm = fitted(bc_glm_countyfixed),
#   fit_gam = fitted(bc_gam_countyfixed)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# 
# plot_grid(glmm_gamm_pt, glmm_gamm_hist)
# 
# length(fitted(bc_gamm_county$gam))
# length(fitted(bc_glmm_county))
# length(unique(fitted(bc_gamm_county$gam)))
# length(unique(fitted(bc_glmm_county)))
# bc_gamm_county$mer
```

```{r Compare GLM and GAM using mer}
# # 1. run GLM and GAM with only year as fixed effect
# bc_glm = glm(bc ~ as.factor(year),
#     data = bc_ai_join)
# summary(bc_glm)
# bc_gam = gamm4(bc ~ s(year),
#     data = bc_ai_join)
# summary(bc_gam)
# 
# ## compare GLM and GAM fitted values
# glm_gam_pt = tibble(
#   fit_glm = fitted(bc_glm),
#   fit_gam = fitted(bc_gam$mer)) %>% 
#   ggplot(aes(x = fit_glm, y = fit_gam)) + 
#   geom_point() 
# 
# glm_gam_hist = tibble(
#   fit_glm = fitted(bc_glm),
#   fit_gam = fitted(bc_gam$mer)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# require(cowplot)
# plot_grid(glm_gam_pt, glm_gam_hist, ncol=2)
# ### notes: seem to fit along the identity line (at least, not the horizontal striations we saw above). small range of differences in values based on histogram.
# 
# # 2. run GLM and GAM with year AND county as fixed effects
# ## sample 10,000 rows because otherwise it will take forever to run
# 
# bc_ai_sample = sample_n(bc_ai_join, 10000)
# bc_glm_countyfixed = glm(bc ~ as.factor(year) + County,
#     data = bc_ai_sample)
# summary(bc_glm_countyfixed)
# bc_gam_countyfixed = gamm4(bc ~ s(year) + County,
#     data = bc_ai_sample)
# summary(bc_gam_countyfixed)
# 
# ## compare GLM and GAM fitted values
# glm_gam_county_pt = tibble(
#   fit_glm = predict(bc_glm_countyfixed),
#   fit_gam = predict(bc_gam_countyfixed$mer)) %>% 
#   ggplot(aes(x = fit_glm, y = fit_gam)) + 
#   geom_point() 
# 
# glm_gam_county_hist = tibble(
#   fit_glm = predict(bc_glm_countyfixed),
#   fit_gam = predict(bc_gam_countyfixed$mer)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# 
# 
# plot_grid(glm_gam_county_pt, glm_gam_county_hist)
# fixef(bc_gam_countyfixed$mer)
# ### notes: also seem to fit along the identity line (at least, not the horizontal striations we saw above). small range of differences in values based on histogram. a good sign!
# 
# 
# # 3. Run GLMM and GAMM with county as a nested random effect within states
# ## use sampled 10,000 rows because otherwise it will take forever to run
# bc_glmm_county = lmer(bc ~ as.factor(year) + (1|State/County),
#     data = bc_ai_join)
# summary(bc_glmm_county)
# bc_gamm_county = gamm4(bc ~ s(year, k = 10),
#                            random = ~(1|State/County),
#                        data = bc_ai_join)
# summary(bc_gamm_county$gam)
# 
# ## compare GLMM and GAMM fitted values
# glmm_gamm_pt = tibble(
#   fit_glmm = fitted(bc_glmm_county),
#   fit_gamm = fitted(bc_gamm_county$mer)) %>% 
#   ggplot(aes(x = fit_glmm, y = fit_gamm)) + 
#   geom_point() 
# 
# glmm_gamm_hist = tibble(
#   fit_glm = fitted(bc_glm_countyfixed),
#   fit_gam = fitted(bc_gam_countyfixed)) %>% 
#   mutate(diff = fit_glm - fit_gam) %>% 
#   ggplot(aes(x = diff)) + geom_histogram() 
# 
# plot_grid(glmm_gamm_pt, glmm_gamm_hist)
# 
# length(fitted(bc_gamm_county$gam))
# length(fitted(bc_glmm_county))
# length(unique(fitted(bc_gamm_county$gam)))
# length(unique(fitted(bc_glmm_county)))
# bc_gamm_county$mer
```
