---
title: "Additional sensitivity analyses: spatial models"
author: "Maggie Li (ml4424)"
date: "7/7/2021"
output: html_document
---

## Read in packages

```{r}
library(knitr)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
```

```{r load packages}
library(spdep)
library(rgdal)
library(rgeos)
library(tidyverse)
library(tidycensus)
library(splm)
library(tmap)
library(lme4)
library(gamm4)
```

```{r Read in data}
# read in files, wrangle data to join with spatial dataset
bc = read_csv("../data/county_concentrations/joined_dta/bc_ai_join.csv") %>%
  rename(GEOID = County) %>%
  arrange((as.numeric(GEOID))) %>%
  dplyr::select(GEOID, bc, State, year, county_type, popd_q, hhinc_q, Climate_Zone)
```

# GAMM tensor term for spatial autocorrelation

## Get county shapefile and latitude longitude of county centroids

```{r cache = TRUE}
# get county centroid lat/lon coordinates for tensor smooth product term
# census_api_key("ef9bc9f1392567620517b2f5ba86c86ebbd70d69", overwrite=TRUE, install = TRUE)
# readRenviron("~/.Renviron")
county <- get_decennial(
  geography = "county",
  variables = c(county = "COUNTY", state = "STATE"),
  output = "wide",
  geometry = TRUE
  ) %>% 
  filter(state != "2" & state != "15" & state <= 56) %>% # filter out AL and HI, and territories
  mutate(GEOID = str_replace(GEOID, "46113", "46102")) %>% 
  filter(GEOID != "51515")

# compare w/ us_counties df
# county$GEOID
# us_counties$GEOID
# setdiff(county$GEOID, us_counties$GEOID) #compare these vectors
# setdiff(us_counties$GEOID, county$GEOID) 

# NOTES: Oglala County (46102) was renamed from Shannon County (46113) in 2015; Bedford City (51515) joined Bedford County in 2013 (51019); so use str_replace to change to 46102 and drop 51515 FIPS

count_centroid <- county %>% 
  st_centroid()
# count_centroid
# st_coordinates(count_centroid$geometry) # this function extracts lat long but into a vector, not what we want

count_centroid <- count_centroid %>% mutate(long = unlist(map(count_centroid$geometry,1)),
           lat = unlist(map(count_centroid$geometry,2))) # extract centroid lat long from sf into the same df. yay!

# Join county lat lon data with bc
bc_lat_lon = left_join(bc, count_centroid)

bc_lat_lon
```

## Main effects only (no county type x year interX term)

#### 10 knots
```{r cache = TRUE}
# bc_tensor_mod_10 =  gamm4(bc ~ county_type + as.factor(year) + popd_q + hhinc_q
#                         + t2(lat, long,
#                              k = c(10, 10), # try default 10 knots
#                              bs = 'cr'),
#                  random = ~(1|State/GEOID),
#                  data = bc_lat_lon) 
# 
# summary(bc_tensor_mod_10$gam)
# summary(bc_tensor_mod_10$mer)
# bc_tensor_mod_10$gam
# 
# ## save model output as RDS to read back in later
# saveRDS(bc_tensor_mod_10,
#         file = "../intermediate/model_outputs/spatial/main_effects_only/bc_tensor_10knots.rds")
bc_tensor_mod_10 = readRDS("../intermediate/model_outputs/spatial/main_effects_only/bc_tensor_10knots.rds")
bc_tensor_mod_10
summary(bc_tensor_mod_10$mer)
summary(bc_tensor_mod_10$mer)$coefficients[2,2]

## Extract 95% CI: point estimate, LL, UL
paste(summary(bc_tensor_mod_10$mer)$coefficients[2,1] %>% round(digits = 3),
      " (",
      round(summary(bc_tensor_mod_10$mer)$coefficients[2,1] - 1.96*summary(bc_tensor_mod_10$mer)$coefficients[2,2], digits = 3),
      ", ",
      round(summary(bc_tensor_mod_10$mer)$coefficients[2,1] + 1.96*summary(bc_tensor_mod_10$mer)$coefficients[2,2], digits = 3),
      ")", sep = "")
paste(summary(bc_tensor_mod_10$mer)$coefficients[2,1] %>% round(digits = 2),
      " (",
      round(summary(bc_tensor_mod_10$mer)$coefficients[2,1] - 1.96*summary(bc_tensor_mod_10$mer)$coefficients[2,2], digits = 2),
      ", ",
      round(summary(bc_tensor_mod_10$mer)$coefficients[2,1] + 1.96*summary(bc_tensor_mod_10$mer)$coefficients[2,2], digits = 2),
      ")", sep = "")
```

## With county type x year interX term
#### 5 knots
```{r}
# Run GAMM with tensor smooth term
# 
# bc_tensor_mod_5 =  gamm4(bc ~ county_type*as.factor(year) + popd_q + hhinc_q
#                         + t2(lat, long,
#                              k = c(5, 5),
#                              bs = 'cr'),
#                  random = ~(1|State/GEOID),
#                  data = bc_lat_lon) # try 5 knots first
# 
# summary(bc_tensor_mod_5$gam)
# summary(bc_tensor_mod_5$mer)
# 
# # save model output as RDS to read back in later
# saveRDS(bc_tensor_mod_5,
#         file = "../intermediate/model_outputs/spatial/bc_tensor_5knots.rds")
# bc_tensor_mod_5 = readRDS("../intermediate/model_outputs/spatial/bc_tensor_5knots.rds")
# bc_tensor_mod_5
# summary(bc_tensor_mod_5$gam)
```

###### GAMM4 t2 Difference in time trends plot
```{r}
# # vcov matrix
# native_yr_vcov <- vcov(bc_tensor_mod_5$gam)[c(2,seq(38,54)), c(2,seq(38,54))]
# # calculate all the variances for all the years i.e. var(x+y); should be 18 total entries
# var_vector = c()
# for (i in 2:18){
#   var_vector[1] <- native_yr_vcov[1,1]
#   var_vector[i] <- native_yr_vcov[1,1] + native_yr_vcov[i,i] + 2 * native_yr_vcov[i,1]
# }
# native_yr_vcov[1,1] + native_yr_vcov[2,2] + 2 * native_yr_vcov[2,1]
# 
# length(unique(native_yr_vcov))
# sd_vector <- sqrt(var_vector)
# 
# #matrix with 19 cols for 19 years, three rows: one for effect estimate of total effect per year (total effect = main effect + interx effect), one for CI lower, one for CI upper
# bc_decline_model_all <- data.frame()
# bc_decline_model_all[1,1] <- summary(bc_tensor_mod_5$gam)$p.coeff[2] # county type main effect estimate (baseline year)
# bc_decline_model_all[1,2] <- summary(bc_tensor_mod_5$gam)$p.coeff[2] - 1.96*sd_vector[1] # LL 95% CI
# bc_decline_model_all[1,3] <- summary(bc_tensor_mod_5$gam)$p.coeff[2] + 1.96*sd_vector[1] # UL 95% CI
# 
# # fill in matrix thru loop for every following year
# yr_ct <- 38
# for (i in 2:18){
#   bc_decline_model_all[i,1] <- summary(bc_tensor_mod_5$gam)$p.coeff[2]+
#     summary(bc_tensor_mod_5$gam)$p.coeff[yr_ct]
#   bc_decline_model_all[i,2] <- summary(bc_tensor_mod_5$gam)$p.coeff[2]+
#     summary(bc_tensor_mod_5$gam)$p.coeff[yr_ct] - 1.96*sd_vector[i]
#   bc_decline_model_all[i,3] <- summary(bc_tensor_mod_5$gam)$p.coeff[2]+
#     summary(bc_tensor_mod_5$gam)$p.coeff[yr_ct] + 1.96*sd_vector[i]
#   yr_ct <- yr_ct + 1
# }
# # set col names
# colnames(bc_decline_model_all) <- c('estimate', 'cl_lower', 'cl_upper')
# bc_decline_model_all$year <- seq(2000, 2017)
# 
# #PLOT OF TOTAL EFFECT OF NATIVE OVER TIME
# modelsp_interx_plot <-ggplot() + 
#   theme_linedraw() + 
#   geom_line(data = bc_decline_model_all,
#             aes(x=year, y = estimate)) +
#   geom_line(data=bc_decline_model_all,
#             aes(x=year, y=cl_lower), linetype = "dashed") +
#     geom_line(data=bc_decline_model_all,
#             aes(x=year, y=cl_upper), linetype = "dashed") +
#   ylim(-0.5, 0.5) +
#   # ggtitle(expression(paste("Modeled ", PM[2.5], " Difference in AI vs. Non-AI Populated Counties"))) +
#   ylab(expression(paste("Mean Difference in BC (", mu, "g/", m^3, ")"))) +
#   # xlab("Year") +
#   theme(plot.title = element_text(size = 18),
#         axis.title = element_text(size = 16),
#         axis.text = element_text(size = 12),
#         # axis.title.x = element_blank(),
#         # axis.title.y = element_blank()
#         ) +
#   guides(x =  guide_axis(angle = 45)) +
#   scale_x_continuous(breaks = seq(2000,2018,1), expand = c(0, 0)) + 
#   geom_hline(yintercept=0, linetype="solid", color = "red") 
# modelsp_interx_plot
# ggsave("../figures/effectmod_plots/sensitivity/spatial/bc_interx_t2_5knots.png")
```


#### 10 knots
```{r cache = TRUE}
# bc_tensor_mod_10 =  gamm4(bc ~ county_type*as.factor(year) + popd_q + hhinc_q
#                         + t2(lat, long,
#                              k = c(10, 10),
#                              bs = 'cr'),
#                  random = ~(1|State/GEOID),
#                  data = bc_lat_lon) # try default 10 knots
# 
# summary(bc_tensor_mod_10$gam)
# summary(bc_tensor_mod_10$mer)
# bc_tensor_mod_10$gam
# 
# # save model output as RDS to read back in later
# saveRDS(bc_tensor_mod_10,
#         file = "../intermediate/model_outputs/spatial/bc_tensor_10knots.rds")
bc_tensor_mod_10 = readRDS("../intermediate/model_outputs/spatial/bc_tensor_10knots.rds")
bc_tensor_mod_10
summary(bc_tensor_mod_10$gam)

```

###### GAMM4 t2 Difference in time trends plot
```{r}
# vcov matrix
native_yr_vcov <- vcov(bc_tensor_mod_10$gam)[c(2,seq(38,54)), c(2,seq(38,54))]
# calculate all the variances for all the years i.e. var(x+y); should be 18 total entries
var_vector = c()
for (i in 2:18){
  var_vector[1] <- native_yr_vcov[1,1]
  var_vector[i] <- native_yr_vcov[1,1] + native_yr_vcov[i,i] + 2 * native_yr_vcov[i,1]
}
native_yr_vcov[1,1] + native_yr_vcov[2,2] + 2 * native_yr_vcov[2,1]

length(unique(native_yr_vcov))
sd_vector <- sqrt(var_vector)

#matrix with 19 cols for 19 years, three rows: one for effect estimate of total effect per year (total effect = main effect + interx effect), one for CI lower, one for CI upper
bc_decline_model_all <- data.frame()
bc_decline_model_all[1,1] <- summary(bc_tensor_mod_10$gam)$p.coeff[2] # county type main effect estimate (baseline year)
bc_decline_model_all[1,2] <- summary(bc_tensor_mod_10$gam)$p.coeff[2] - 1.96*sd_vector[1] # LL 95% CI
bc_decline_model_all[1,3] <- summary(bc_tensor_mod_10$gam)$p.coeff[2] + 1.96*sd_vector[1] # UL 95% CI

# fill in matrix thru loop for every following year
yr_ct <- 38
for (i in 2:18){
  bc_decline_model_all[i,1] <- summary(bc_tensor_mod_10$gam)$p.coeff[2]+
    summary(bc_tensor_mod_10$gam)$p.coeff[yr_ct]
  bc_decline_model_all[i,2] <- summary(bc_tensor_mod_10$gam)$p.coeff[2]+
    summary(bc_tensor_mod_10$gam)$p.coeff[yr_ct] - 1.96*sd_vector[i]
  bc_decline_model_all[i,3] <- summary(bc_tensor_mod_10$gam)$p.coeff[2]+
    summary(bc_tensor_mod_10$gam)$p.coeff[yr_ct] + 1.96*sd_vector[i]
  yr_ct <- yr_ct + 1
}
# set col names
colnames(bc_decline_model_all) <- c('estimate', 'cl_lower', 'cl_upper')
bc_decline_model_all$year <- seq(2000, 2017)

#PLOT OF TOTAL EFFECT OF NATIVE OVER TIME
modelsp_interx_plot <-ggplot() + 
  theme_linedraw() + 
  geom_line(data = bc_decline_model_all,
            aes(x=year, y = estimate)) +
  geom_line(data=bc_decline_model_all,
            aes(x=year, y=cl_lower), linetype = "dashed") +
    geom_line(data=bc_decline_model_all,
            aes(x=year, y=cl_upper), linetype = "dashed") +
  ylim(-0.5, 0.5) +
  # ggtitle(expression(paste("Modeled ", PM[2.5], " Difference in AI vs. Non-AI Populated Counties"))) +
  ylab(expression(paste("Mean Difference in BC (", mu, "g/", m^3, ")"))) +
  xlab("Year") +
  theme(plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 12),
        axis.title.x = element_blank(),
        # axis.title.y = element_blank()
        ) +
  guides(x =  guide_axis(angle = 45)) +
  scale_x_continuous(breaks = seq(2000,2018,1), expand = c(0, 0)) + 
  geom_hline(yintercept=0, linetype="solid", color = "red") 
modelsp_interx_plot
ggsave("../figures/effectmod_plots/sensitivity/spatial/bc_interx_t2_10knots.png")
```


##### set tensor term to 20 knots for lat and lon

```{r cache = TRUE}
# gc()
# bc_tensor_mod_20 =  gamm4(bc ~ county_type*as.factor(year) + popd_q + hhinc_q
#                         + t2(lat, long, 
#                              k = c(20, 20),
#                              bs = 'cr'),
#                  random = ~(1|State/GEOID),
#                  data = bc_lat_lon) # try 20 knots 
# 
# summary(bc_tensor_mod_20$gam)
# summary(bc_tensor_mod_20$mer)
# bc_tensor_mod_20$gam
# 
# 
# ## save model output as RDS to read back in later
# saveRDS(bc_tensor_mod_20,
#         file = "../intermediate/model_outputs/spatial/bc_tensor_mod_20.rds")
bc_tensor_mod_20 = readRDS("../intermediate/model_outputs/spatial/bc_tensor_mod_20.rds")
bc_tensor_mod_20
summary(bc_tensor_mod_20$gam)
```
###### GAMM4 t2 Difference in time trends plot: 20 knots

```{r}
# vcov matrix
native_yr_vcov <- vcov(bc_tensor_mod_20$gam)[c(2,seq(38,54)), c(2,seq(38,54))]
# calculate all the variances for all the years i.e. var(x+y); should be 18 total entries
var_vector = c()
for (i in 2:18){
  var_vector[1] <- native_yr_vcov[1,1]
  var_vector[i] <- native_yr_vcov[1,1] + native_yr_vcov[i,i] + 2 * native_yr_vcov[i,1]
}
native_yr_vcov[1,1] + native_yr_vcov[2,2] + 2 * native_yr_vcov[2,1]

length(unique(native_yr_vcov))
sd_vector <- sqrt(var_vector)

#matrix with 19 cols for 19 years, three rows: one for effect estimate of total effect per year (total effect = main effect + interx effect), one for CI lower, one for CI upper
bc_decline_model_all <- data.frame()
bc_decline_model_all[1,1] <- summary(bc_tensor_mod_20$gam)$p.coeff[2] # county type main effect estimate (baseline year)
bc_decline_model_all[1,2] <- summary(bc_tensor_mod_20$gam)$p.coeff[2] - 1.96*sd_vector[1] # LL 95% CI
bc_decline_model_all[1,3] <- summary(bc_tensor_mod_20$gam)$p.coeff[2] + 1.96*sd_vector[1] # UL 95% CI

# fill in matrix thru loop for every following year
yr_ct <- 38
for (i in 2:18){
  bc_decline_model_all[i,1] <- summary(bc_tensor_mod_20$gam)$p.coeff[2]+
    summary(bc_tensor_mod_20$gam)$p.coeff[yr_ct]
  bc_decline_model_all[i,2] <- summary(bc_tensor_mod_20$gam)$p.coeff[2]+
    summary(bc_tensor_mod_20$gam)$p.coeff[yr_ct] - 1.96*sd_vector[i]
  bc_decline_model_all[i,3] <- summary(bc_tensor_mod_20$gam)$p.coeff[2]+
    summary(bc_tensor_mod_20$gam)$p.coeff[yr_ct] + 1.96*sd_vector[i]
  yr_ct <- yr_ct + 1
}
# set col names
colnames(bc_decline_model_all) <- c('estimate', 'cl_lower', 'cl_upper')
bc_decline_model_all$year <- seq(2000, 2017)

#PLOT OF TOTAL EFFECT OF NATIVE OVER TIME
modelsp_interx_plot <-ggplot() + 
  theme_linedraw() + 
  geom_line(data = bc_decline_model_all,
            aes(x=year, y = estimate)) +
  geom_line(data=bc_decline_model_all,
            aes(x=year, y=cl_lower), linetype = "dashed") +
    geom_line(data=bc_decline_model_all,
            aes(x=year, y=cl_upper), linetype = "dashed") +
  ylim(-0.5, 0.5) +
  # ggtitle(expression(paste("Modeled ", PM[2.5], " Difference in AI vs. Non-AI Populated Counties"))) +
  ylab(expression(paste("Mean Difference in BC (", mu, "g/", m^3, ")"))) +
  xlab("Year") +
  theme(plot.title = element_text(size = 18),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 12),
        # axis.title.x = element_blank(),
        # axis.title.y = element_blank()
        ) +
  guides(x =  guide_axis(angle = 45)) +
  scale_x_continuous(breaks = seq(2000,2018,1), expand = c(0, 0)) + 
  geom_hline(yintercept=0, linetype="solid", color = "red") 
modelsp_interx_plot
ggsave("../figures/effectmod_plots/sensitivity/spatial/bc_interx_t2_20.png")
```